\NAME: A framework for \\ Efficient Intrusion Detection\\ in the Internet of Things

### 2014-10-17 paper.tex
clean up of classification of ID algorithms

\documentclass[conference]{IEEEtran}
\usepackage[pdftex]{graphicx}
\usepackage[cmex10]{amsmath}
\interdisplaylinepenalty=2500
% \usepackage{algorithmic}
\usepackage{array}
\usepackage{mdwmath}
\usepackage{mdwtab}
\usepackage{eqparbox}
\usepackage[font=footnotesize]{subfig}
\usepackage{fixltx2e}
\usepackage{url}
\hyphenation{}

% additional packages and utility commands
\newcommand{\TODO}{\textbf{\color{red}TODO}}

\usepackage{minted}
\usepackage{flushend}

@@

\title{
The \NAME Framework:\\
Efficient Intrusion Detection\\
in the Internet of Things
}

@@

The IoT presents a great potential to positively influence our daily work and
life: domotics, assisted living, e-health and enhanced learning are only a few
examples\cite{atzori2010internet}. But with this potential the IoT also
presents a significant threat: by opening our smart homes
\cite{aldrich2003smart} and personal data\cite{weber2010internet} to all these
interconnected devices, we also open them to everyone who is able to break
these new, virtual locks. IoT devices typically have limited batteries, little
processing power and memory is restricted to the bare minimum. These properties
make it hard to add any additional service, including security.

@@

In wireless networks of resource constraint devices, which make up an important
part of the IoT, it is not possible to have this single point in the network
that can oversee all traffic \cite{zhang2000intrusion, mishra2004intrusion}. If
the outer security perimeter diminishes, every device has to implement its own
lines of defense. Making the IDS a local service on each device requires
resources, which are already very constraint.

@@

Current developments in ID focus on programming frameworks to structure the
implementation of different algorithms \cite{valero2012di} and offer the
required basic functional components to implement actual algorithms
\cite{krontiris2008lidea}, but don't offer a way to optimize the usage of
resources. They also don't support systematic reuse to create different
configurations for different nodes or evolution in time.

@@

The IoT is a heterogeneous environment, both in hardware and software.
Addressing this in a transparent and automated way is a key success factor for
any kind of development in this area. A solution that supports the integration
of multiple algorithms on IoT devices needs to consider the impact on their
resources. Different algorithms should be merged to avoid accumulating their
impact.

@@

The first contribution of this paper is a framework that enables the formal
description of ID algorithms on a functional level. An automated code
generation process produces corresponding source code, for a given platform and
configuration. This way, the framework addresses the heterogeneous nature of
both the IoT, as a hardware and software platform, and ID algorithms.

@@

The second contribution introduces functional code fusion (FCF) as a generation
paradigm. FCF identifies common data and functions, and organizes code in such
a way that redundant iterations, tests and computations are eliminated. Our
results show substantially improved memory usage, execution time and usage of
the wireless radio, reducing energy consumption.

@@

The remainder of this paper proceeds as follows: section \ref{landscape} takes
a look at ID algorithms and classifies them. Section \ref{analysis} looks for
patterns in the implementation of these classes of algorithms, details some of
the ideas behind FCF and identifies the different aspects of the problem space.
Section \ref{design} describes the design we applied to construct the
framework, its domain specific language (DSL) and code generator. Section
\ref{evaluation} evaluates an implementation and application of the framework.
Section \ref{related} explores related work in the field of DSLs for ID and
code generation for the IoT. Finally, section \ref{conclusion} summarizes our
findings, draws conclusions and identifies topics for future work.

\section{Landscape of ID Algorithms}
\label{landscape}

@@

\subsection{Detecting Intrusions}
\label{subsection:detecting}

@@

The meshed network topology of WSNs and its routing protocols have produced
many related attacks and each of these attacks has been the focus of many
research papers presenting algorithms to detect them. In the following
paragraphs we take a look at some popular ones.

In a wormhole attack, an artificial low-latency link is created between two
distant nodes in the network. This causes the routing protocol to divert more
traffic through this link, offering the attacker an abundance of packets to
inspect. In \cite{maheshwari2007detecting} an algorithm is proposed to allow
nodes to determine if a wormhole is actively present. In essence, the algorithm
relies on the exchange of neighbour lists to allow nodes to determine if
unexpected substructures are present in the graph representing the network.

Another attack that is related to the wormhole is the sinkhole. When launching
a sinkhole attack, as described in \cite{krontiris2008launching}, the attacker
first needs to compromise an existing node. Being in control of the node, the
attacker can make his captured node look more \emph{attractive} to the
surrounding nodes and thus draw more traffic to itself. The sinkhole attack is
typically a foundation for other routing level attacks, such as selective
forwarding, modification of packets or even dropping them altogether. In its
simplest form, the sinkhole offers the attacker with lots of packets to analyze
and use to gather information about the network and its functionality.

Detecting sinkholes is very hard and is often only possible based on other
attacks that are supported by the sinkhole. When the sinkhole is used to
implement selective forwarding, \cite{ngai2006intruder} proposes an algorithm
that allows a base station, the aggregating master node to which all nodes
typically send their information, to observe missing data from nodes in the
same area in a statistical way.

More typical attacks such as flooding, the sybil attack, rushing\dots are
amongst others presented in \cite{wood2002denial} and \cite{djenouri2005survey}.

\subsection{Cooperative Decision Making}
\label{subsection:coorperative}

As in many other situations, a group is often stronger than the sum of its
individual components. This is surely the case for WSNs. Because not all nodes
are always actively participating in the network, some nodes might miss clues
that would otherwise lead them to detect intrusions. It is hardly impossible
for a single node to detect a complex attack by itself. Therefore a second
important research topic consists of cooperative algorithms to combine
information from single nodes into a group-based decision about alleged
intrusions.

In \cite{krontiris2009cooperative}, the authors first present a theoretical
foundation to analyze cooperative algorithms. Given these foundations they
continue to present an algorithm consisting of 5 phases. It essentially
implements a voting system based on the Guy Fawkes protocol
\cite{anderson1998new} to identify an intruder. It enables the exchange of
suspected intruder information and allows distributed authentication of those
\emph{votes}. Based on these authenticated votes, a distributed decision can be
made about the commonly identified intruder.

A distributed, cooperative algorithm can sometimes be implemented locally. This
is illustrated by the reputation-based detection algorithm introduced in
\cite{ganeriwal2008reputation}. Using this algorithm, nodes can exchange
information about the reputation of other nodes. Combining this information
allows them to decide about their trust in a given node, based on more than
their own, often partial, observations.

@@

A research topic that exists in parallel to the detection and cooperation
duality is software attestation. Software attestation offers algorithms for
exchanging information about the software that is running on a node, and aims
to identify nodes that can no longer prove that their content is unaltered.
Examples of evolving algorithms to implement this functionality include SWATT
\cite{seshadri2004swatt}, ICE/SCUBA \cite{seshadri2006scuba} and SAKE
\cite{seshadri2008sake}.

@@

\section{Analysis of ID Algorithms}
\label{analysis}

>>>>

\documentclass[conference]{IEEEtran}
\usepackage[pdftex]{graphicx}
\usepackage[cmex10]{amsmath}
\interdisplaylinepenalty=2500
\usepackage{array}
\usepackage{mdwmath}
\usepackage{mdwtab}
\usepackage{eqparbox}
\usepackage[font=footnotesize]{subfig}
\usepackage{fixltx2e}
\usepackage{url}
\hyphenation{}

% additional packages and utility commands

\newcommand{\TODO}{\textbf{\color{red}TODO}}

\usepackage{flushend}

@@

\title{
\NAME: A framework for \\
Efficient Intrusion Detection\\
in the Internet of Things
}

@@

The IoT presents a great potential to positively influence our daily work and
life: domotics, assisted living, e-health and enhanced learning are only a few
examples\cite{atzori2010internet}. But with this potential the IoT also
presents a significant threat: by opening our smart homes
\cite{aldrich2003smart} and personal data\cite{weber2010internet} to all these
interconnected devices, we also open them to everyone who is able to break the
virtual locks that protect them. IoT devices typically have limited batteries,
little processing power and memory is restricted to the bare minimum. These
properties make it hard to add any additional service, including security.

@@

In wireless networks of resource constraint devices, which make up an
significant part of the IoT, it is not possible to have this single point in
the network that can oversee all traffic\cite{zhang2000intrusion,
mishra2004intrusion}. If the outer security perimeter diminishes, every device
has to implement its own lines of defense. But making the IDS a local service
on each device requires local resources, which are the scarce commodity to IoT
devices.

@@

Current developments in ID focus on programming frameworks to structure the
implementation of algorithms\cite{valero2012di} and offer the required basic
functional components to implement them\cite{krontiris2008lidea}. But they
don't offer a way to optimize the usage of resources. Nor do they support
systematic reuse to create different configurations for different devices.

@@

The IoT is a heterogeneous environment, both in hardware and software.
Addressing this in a transparent and automated way is a key success factor for
any kind of development in this area. A solution that supports the integration
of multiple algorithms on IoT devices needs to consider the impact on their
resources. Different algorithms should be merged somehow to avoid accumulating
their impact.

@@

The first contribution of this paper is a framework that enables the formal
description of ID algorithms on a functional level. An automated code
generation process produces corresponding source code, for a given platform and
configuration. This way, the framework addresses the heterogeneous nature of
both the IoT and ID algorithms.

@@

The second contribution introduces functional code fusion (FCF) as a generation
paradigm to solve the problem of hard to reuse and combine ID algorithms. FCF
identifies common data and functions, and organizes code in such a way that
redundant iterations, tests and computations are eliminated. Our results show
substantially improved memory usage, execution time and usage of the wireless
radio, reducing energy consumption.

@@

The remainder of this paper proceeds as follows: section \ref{classification}
takes a look at ID algorithms and classifies them. Section \ref{pattern} looks
for patterns in the implementation of these classes of algorithms, details some
of the ideas behind FCF and identifies the different aspects of the problem
space. Section \ref{design} describes the design we applied to construct the
framework, its domain specific language (DSL) and code generator. Section
\ref{evaluation} evaluates an implementation and application of the framework.
Section \ref{related} explores related work in the field of DSLs for ID and
code generation for the IoT. Finally, section \ref{conclusion} summarizes our
findings, draws conclusions and identifies topics for future work.

\section{A Classification of ID Algorithms}
\label{classification}

@@

\subsection{Detecting Intrusions}
\label{detection}

@@

\subsection{Cooperative Decision Making}
\label{coorperative}

These taxonomies focus on the act of detecting intrusions itself, the local
algorithm so to speak. In a highly distributed environment, the orthogonal
categorization, dealing with the distributed algorithm, is maybe even more
important: how do algorithms deal with the information coming from the group
they are part of?

As in many other situations, a group is often stronger than the sum of its
individual components. Because not all devices are always actively
participating in the network, some might miss clues that would otherwise lead
them to detect intrusions. It is hardly impossible for a single device to
detect a complex attack by itself. Therefore a second important research topic
consists of cooperative algorithms to combine information from single
identities into a group-based decision about alleged
intrusions\cite{zhang2000intrusion}.

A distributed, cooperative algorithm can sometimes be implemented locally. This
is illustrated by the reputation-based detection algorithm introduced in
\cite{ganeriwal2008reputation}, based on the same principle of a watchdog also
found in \cite{mishra2004intrusion} Using this algorithm, nodes can exchange
information about the reputation of other nodes. Combining this information
allows them to decide about their trust in a given node, based on more than
their own, often partial, observations alone.

In \cite{krontiris2009cooperative}, the authors first present a theoretical
foundation to analyze cooperative algorithms. Given these foundations they
continue to present an algorithm consisting of 5 phases. It essentially
implements a voting system based on the Guy Fawkes
protocol\cite{anderson1998new} to identify an intruder. It enables the exchange
of suspected intruder information and allows distributed authentication of
those \emph{votes}. Based on these authenticated votes, a distributed decision
can be made about the commonly identified intruder.

@@

A research topic that exists in parallel to the detection and cooperation
duality is software attestation. Software attestation offers algorithms for
exchanging information about the software that is running on a node, and aims
to identify nodes that can no longer prove that their content is unaltered.
Examples of a family of algorithms to implement this functionality include
SWATT\cite{seshadri2004swatt}, ICE/SCUBA\cite{seshadri2006scuba} and
SAKE\cite{seshadri2008sake}.

@@

\section{A Design Pattern for ID Algorithms}
\label{pattern}
